---
title: "ANOVA example"
output: html_notebook
---
```{r setup}
if (!require("pacman")) install.packages("pacman", repos='https://stat.ethz.ch/CRAN/'); library(pacman)
p_load(ggplot2, 
       DT, 
       plyr,
       dplyr, 
       tidyr,
       assertthat,
       car,
       GetoptLong,
       tibble,
       cowplot,
       readr,
       broom,
       GGally)

# string interpolation
qq.options("code.pattern" = "#\\{CODE\\}") 

# plot theme
myTheme <- theme(panel.background = element_blank(), panel.grid.major = element_line(color="lightgrey", size = 0.2))

# decimal output
options(digits=2)

# datatable
options(DT.options = list(pageLength = 10))
options(DT.autoHideNavigation = TRUE)

# We ensure that the following packages are installed, but left unloaded to 
# prevent conflicts.
# (We can still use the functions inside, e.g., MASS::stepAIC() )

p_load(MASS); p_unload(MASS)     # for stepwise regression 
                
```

# Data
This dataset is simulated from the following hypothetical experiment:

> A company has recently invested in smart watches (known as the Mango) for all sales executives. They have tasked your team to evaluate the impact of the Mango on productivity among the sales staff. They believe that having email on one’s wrist will cause faster response times to customers, and therefore more sales. 
> Data will be collected for Sales Team A over a period of 3 months after the introduction of Mangos. Control data will be collected from members of Sales Team B over the same period—this group will not yet be equipped with Mangos. However, they will receive their Mangos in month 4, giving the researchers both between subjects (A compared to B) and within subjects (pre-Mango B compared to post-Mango B) data.

This analysis will focus on the relationship between inattention score (Kushlev et al. (2016)) and the usage time of the first month (in minutes). We hypothesize that users with higher inattention score will use the watch longer.

```{r}
age_df          <- read_csv("../data/Mango Watch/age.csv")
usageTime_df    <- read_csv("../data/Mango Watch/usageTime.csv")
team_df         <- read_csv("../data/Mango Watch/team.csv")
sus_df          <- read_csv("../data/Mango Watch/SUS.csv")
responseTime_df <- read_csv("../data/Mango Watch/responseTime.csv")
inattention_df  <- read_csv("../data/Mango Watch/inattention.csv")

responseTime_df <- left_join(team_df, responseTime_df, by = "ID")

ageUsageTime_df <- left_join(team_df, age_df, by = "ID") %>% 
  left_join(usageTime_df, by = "ID") %>% 
  left_join(inattention_df, by = "ID")

sus_df <- left_join(team_df, sus_df, by = "ID")
```


# Exploratory plot
```{r}
time_inattention <- ageUsageTime_df %>% 
  ggplot(., aes(x = Inattention, y = UsageTime)) +
  geom_point() +
  # geom_smooth(method = "loess", size = 1.5) +
  geom_smooth(method = "lm", size = 1.5, color = "red") +
  xlim(0,4.1)

lm_inattention <- lm(UsageTime ~ Inattention, ageUsageTime_df)
p_TIresid <- ageUsageTime_df %>% 
  ggplot(., aes(x = Inattention, y = lm_inattention$residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, col = "red") +
  xlim(0,4.1)

plot_grid(time_inattention, p_TIresid, ncol = 2)
```

# Linear model
```{r}
lm1 <- lm(UsageTime ~ Inattention, ageUsageTime_df)
summary(lm1)
summ <- summary(lm1)
lm1_t <- tidy(lm1)
lm_ci <- confint(lm1, level = 0.95)
p.value <- pf(summ$fstatistic[1], summ$fstatistic[2], summ$fstatistic[3], lower.tail = FALSE)
```


__Interpretation:__ A simple linear regression was calculated to predict usage time based on inattention score. A statistically significant regression equation was found (F(`r summ$fstatistic[2]`, `r summ$fstatistic[3]`) = `r summ$fstatistic[1]`, p < .001*) with an R^2 = `r summ$r.squared`.

The predicted usage time is equal to `r lm1_t$estimate[1]` + `r lm1_t$estimate[2]` × (inattention score) minutes when the inattention score is measured according to Kushlev et al. (2016).

The usage time increased `r lm1_t$estimate[2]` minutes for each point of the inattention score (95% CI [`r lm_ci["Inattention", "2.5 %"]`, `r lm_ci["Inattention", "97.5 %"]`]).

* exact p-value: `r p.value`


# Analysis of variance (ANOVA)
Comparing to the null model: (there's no relationship between inattention and usage time)
```{r}
anova_r <- anova(lm_inattention)
print(anova_r)
anova_t <- tidy(anova_r)
```

__Interpretation:__ ANOVA indicates that the model `UsageTime ~ Inattention` score fits the data better than the null model `UsageTime ~ 1`, *F*(`r anova_t$df[1]`, `r anova_t$df[2]`) = `r anova_t$statistic[1]`, p < .001 (exact p = `r anova_t$p.value[1]`).

Alternatively, we can specify the null model explicitly. This would yield the same results
```{r}
anova(lm(UsageTime ~ 1, ageUsageTime_df),             # null model (only intercept)
      lm(UsageTime ~ Inattention, ageUsageTime_df)    # intercept and inattention 
)
```


We can add the team into the model and compare them
```{r}
anova_r <-
  anova(lm(UsageTime ~ 1, ageUsageTime_df),                   # null model (only intercept)
        lm(UsageTime ~ Inattention, ageUsageTime_df),         # intercept and inattention 
        lm(UsageTime ~ Inattention + Team, ageUsageTime_df)   # intercept, inattention, and team
  )
print(anova_r)
# alternative way to modify the model
# lm_inattention_team <- update(lm_inattention, . ~ Team ) # adds a term to the right hand side
```

__Interpretation:__ Adding team as an additonal factor yielded a better fit compared to the model with inattention alone as shown in the following table:

```{r}
anova_t <- 
  tidy(anova_r) %>% 
  mutate(F = statistic) %>% 
  select(-rss, -sumsq, -statistic) %>% 
  select(df, res.df, F, p.value)
anova_t
```


# Your turn
Try fitting a linear model between age, usage time, and team
```{r}
# TODO: your code here
# ageUsageTime_df$Age
# ageUsageTime_df$UsageTime
# ageUsageTime_df$Team
```


# Determining regressors
In science, we prefer simpler models. This is a principle known as *Occam's razor*. 
F-statistics represents how good the model fits the data, but it doesn't penalize
complex model. Instead, we use *Akaike information criterion (AIC)* (read "a-kai-kay"; Japanese). AIC accounts for both the fit and the number of parameters in the model.
**We prefer model with minimum AIC.**

## How to select variables to be regressors?

In HCI, we usually use linear models to analyze survey responses. The questions in
the survey are usually designed with a theoretical basis. Therefore, we usually 
construct the model by adding factors that are theoretically sound.
```{r}
m1 <- lm(UsageTime ~ 1, ageUsageTime_df)
m2 <- update(m1, . ~ Inattention)
m3 <- update(m1, . ~ Team)
m4 <- update(m1, . ~ Inattention + Team)

AIC_result <- data.frame(rbind(
  extractAIC(m1),
  extractAIC(m2),
  extractAIC(m3),
  extractAIC(m4)))
colnames(AIC_result) <- c("df", "AIC")

print(AIC_result)
```

**Interpretation:** `Inattention` reduces AIC more than `Team`. Thus, `Inattention` should be prioritized more than `Team`.
However, if the model can afford two parameters, both `Inattention` and `Team` lowers AIC, hence should be included.


# Stepwise automatic modeling
**Only for exploratory analysis (when you have no theoretical basis)**, that we automate the model construction. Below, we enter all factors and perform *step-wise backward regression* which gradually remove the terms to minimize AIC. 
```{r}
wholeModel <- lm(UsageTime ~ Inattention + Team, data = ageUsageTime_df)
step_backward_result <- 
  MASS::stepAIC(wholeModel, 
    direction = "backward", 
    trace = TRUE           # this shows the fit trace so we can track AIC (optional)
    )
```

**Interpretation:** Removing either `Team` or `Inattention` doesn't reduce AIC. Therefore, both terms should still be included in the model.

```{r}
# stepAIC output a model, so you can obtain F statistics using typical anova() 
# and t-statistics of each coefficient with summary()
anova(step_backward_result)
summary(step_backward_result)
```


# Your turn
Determine whether ther interaction term should be included in this model or not.
```{r}
# TODO: your code here
```



# References

* Kostadin Kushlev, Jason Proulx, and Elizabeth W. Dunn. 2016. "Silence Your Phones": Smartphone Notifications Increase Inattention and Hyperactivity Symptoms. In Proceedings of the 2016 CHI Conference on Human Factors in Computing Systems (CHI '16). ACM, New York, NY, USA, 1011-1020. DOI: https://doi.org/10.1145/2858036.2858359
