---
title: "Logistic regression"
output: 
  html_notebook: 
    code_folding: hide
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)

####
# BEGIN: Chat's R header (revision: 22.05.17)
####

# The "pacman" package allows p_load() and p_load_gh() which
# automatically install missing packages and load them
if (!require("pacman")) install.packages("pacman", repos = 'https://stat.ethz.ch/CRAN/'); library(pacman)

p_load(
   car,        # grab bag of useful functions for NHST
   multcomp,   # for glht (planned comparison, post-hoc test)
   lme4,       # for multilevel linear model
   lmerTest,   # provides p-values for lmer()
   tidyverse,  # collection of the tidyverse packages (this automatically load the following):
   #dplyr,     #   - for data wrangling
   #tibble,    #   - a stricter alternative to data.frame
   #readr,     #   - a stricter alternative to read.csv
   #ggplot2,   #   - for plotting
   #tidyr,     #   - for data tidying
   #purrr,     #   - for functional programming
               # other packages in tidyverse that are non-core
   stringr,    #   - for string functions
   forcats,    #   - utility functions for working with factor levels
   broom,      #   - for cleaing output from models, e.g., lm()
   lubridate,  #   - utility for parsing and performing arithematic on dates 
   modelr,     #   - utility functions for interacting with models in pipe
               # extensions of tidyverse
   cowplot,    #   - adds plot_grid() to put multiple ggplot()'s togeter
   GGally,     #   - adds ggpairs() which is a smarter scatterplot matrix
               # 
               # visualization & interactivity
   Hmisc,      #   - for plotting mean and CI in ggplot
   rafalib,    #   - for imagemat function (visualize contrast codings)
   DT,         #   - for showing data table with navigation/search controls
               # testing:
   assertthat  #   - unit-testing functions
)

# p_load_gh(
#    "eclarke/ggbeeswarm" # beeswarm plot extension for ggplot2
# )

# ggplot2 config (plot theme)
myTheme <- theme(
   panel.background = element_blank(), 
   panel.grid.major = element_line(color = "lightgrey", size = 0.2)
)

# DT config
options(DT.options = list(pageLength = 10))
options(DT.autoHideNavigation = TRUE)

# Hints:
# string interpolation: str_interp("${var}")
# options(digits=2)  # Decimal output readability, but may be problematic with rounding (default value = 7)

####
# END: Chat's R header
####

# additional packages
p_load(car) # for Anova with Type II test (see below)
p_load(ggmosaic) # for Mosaic plot
```

Scenario: We asked the participants two questions:

* How frequent do you *read* information on social media?
* How frequent do you *post* information on social media?

The responses are from 1 (very seldom) -- 7 (very frequent)
We wish to know whether the self-perception of social media behavior and their age influences their preference of social media (Facebook vs. Twitter).

Load and visualize the data
```{r}
sns_behavior_df <- read_csv("../data/sns_behavior.csv", col_types = "iiic")

p_scatter <- 
  sns_behavior_df %>% 
  ggplot(aes(x = Reading, y = Posting, shape = App, fill = Age, color = Age)) +
  geom_point(position = position_dodge(width = 0.2))
p_scatter
```

Prepare the dependent variable (setting the baseline level).
```{r}
# TODO
```

# Model 1: Age only

Fit logistic regression model and perform Type II ANOVA.

```{r}
# m_logit <- glm( (formula here) , data = sns_behavior_df, family = binomial(link = "logit"))
# TODO
```

Extract coefficients and calculate confidence interval
```{r}
# m_logit_coeff <- tidy(m_logit, TODO) # some additional parameters needed
```

Plot the data and prediction

```{r}
# TODO
```

Plot the odds of each effect:

```{r}
# TODO plot the odds
```
TODO: Interpret the ANOVA results and the odds results


# Model 2: Age, Reading, and Posting rates

Fit logistic regression model and perform Type II ANOVA.

```{r}
# m_logit <- glm( (formula here) , data = sns_behavior_df, family = binomial(link = "logit"))
# TODO
```

Extract coefficients and calculate confidence interval
```{r}
# m_logit_coeff <- tidy(m_logit, TODO) # some additional parameters needed
```

Plot the data and prediction

```{r}
# TODO
```

Plot the odds of each effect:

```{r}
# TODO plot the odds
```
TODO: Interpret the ANOVA results and the odds results

