---
title: "TODO"
output: 
  html_notebook: 
    code_folding: hide
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)

####
# BEGIN: Chat's R header (revision: 15.04.17)
####

# The "pacman" package allows p_load() and p_load_gh() which
# automatically install missing packages and load them
if (!require("pacman")) install.packages("pacman", repos = 'https://stat.ethz.ch/CRAN/'); library(pacman)

p_load(
   car,        # grab bag of useful functions for NHST
   multcomp,   # for glht (planned comparison, post-hoc test)
   lme4,       # for multilevel linear model
   lmerTest,   # provides p-values for lmer()
   tidyverse,  # collection of the tidyverse packages (this automatically load the following):
   #dplyr,     #   - for data wrangling
   #tibble,    #   - a stricter alternative to data.frame
   #readr,     #   - a stricter alternative to read.csv
   #ggplot2,   #   - for plotting
   #tidyr,     #   - for data tidying
   #purrr,     #   - for functional programming
               # other packages in tidyverse that are non-core
   stringr,    #   - for string functions
   forcats,    #   - utility functions for working with factor levels
   broom,      #   - for cleaing output from models, e.g., lm()
   lubridate,  #   - utility for parsing and performing arithematic on dates 
               # extensions of tidyverse
   cowplot,    #   - adds plot_grid() to put multiple ggplot()'s togeter
   GGally,     #   - adds ggpairs() which is a smarter scatterplot matrix
               # 
               # visualization & interactivity
   Hmisc,      #   - for plotting mean and CI in ggplot
   rafalib,    #   - for imagemat function (visualize contrast codings)
   DT,         #   - for showing data table with navigation/search controls
               # testing:
   assertthat  #   - unit-testing functions
)

p_load_gh(
   "eclarke/ggbeeswarm" # beeswarm plot extension for ggplot2
)

# ggplot2 config (plot theme)
myTheme <- theme(
   panel.background = element_blank(), 
   panel.grid.major = element_line(color = "lightgrey", size = 0.2)
)

# DT config
options(DT.options = list(pageLength = 10))
options(DT.autoHideNavigation = TRUE)

# Hints:
# string interpolation: str_interp("${var}")
# options(digits=2)  # Decimal output readability, but may be problematic with rounding (default value = 7)

####
# END: Chat's R header
####

p_load(
  ARTool,   # for Aligned Rank Transform art()
  nortest, # for lillie.test(): The Kolmogorov-Smirnov test with a correct p-value
  fGarch,  # for rsnorm(): generating skewed distributions
  lawstat  # for Brown-Forsythe test: levene.test(location = "median")
  )
p_unload(ARTool, nortest, fGarch, lawstat)
```

# Data
This dataset is simulated from the following hypothetical experiment:

> A company has recently invested in smart watches (known as the Mango) for all sales executives. They have tasked your team to evaluate the impact of the Mango on productivity among the sales staff. They believe that having email on one’s wrist will cause faster response times to customers, and therefore more sales. 
> Data will be collected for Sales Team A over a period of 3 months after the introduction of Mangos. Control data will be collected from members of Sales Team B over the same period—this group will not yet be equipped with Mangos. However, they will receive their Mangos in month 4, giving the researchers both between subjects (A compared to B) and within subjects (pre-Mango B compared to post-Mango B) data.


```{r}
age_df          <- read_csv("../data/Mango Watch/age.csv")
usageTime_df    <- read_csv("../data/Mango Watch/usageTime.csv")
team_df         <- read_csv("../data/Mango Watch/team.csv")
sus_df          <- read_csv("../data/Mango Watch/SUS.csv")
responseTime_df <- read_csv("../data/Mango Watch/responseTime.csv")
inattention_df  <- read_csv("../data/Mango Watch/inattention.csv")

responseTime_df <- left_join(team_df, responseTime_df, by = "ID")

ageUsageTime_df <- left_join(team_df, age_df, by = "ID") %>% 
  left_join(usageTime_df, by = "ID") %>% 
  left_join(inattention_df, by = "ID")

sus_df <- left_join(team_df, sus_df, by = "ID")
```

Your turn: analyze the response time by team and by month (separately). Don't forget statistical assumptions.

```{r}
responseTime_df %>% 
  ggplot(., aes(x = ResponseTime) ) +
  geom_histogram(bins = 80) +
  facet_grid(Team + Month ~ . )
```


# Normality of residual
```{r}
#### Hints
# normality: plot
# car::qqPlot(diff)

# normality: test
# nortest::lillie.test(diff)
```

# Equal variance

```{r}
#### Hints
# Equal variance: median plot
# data %>% 
#   group_by(x) %>% 
#   summarise(Median = median(y)) %>% 
#   ungroup() %>% 
#   inner_join(data, by = "x") ->
#   data_median
# 
# p_median_diff <-
#   data_median %>% 
#   mutate(yMedDiff = y - Median) %>% 
#   ggplot(aes(x = x, y = yMedDiff)) +
#   geom_boxplot() +
#   geom_hline(yintercept = 0, color = "green") +
#   ylab("y - Median(y)")


# Equal variance: test
# lawstat::levene.test(data$y, data$x, location = "median")
```


# Transformation
```{r}
# Check Log-transformation from Assumptions.Rmd


# Check ART-transformation from AlignedRankTransform.Rmd
# ARTool::art(Y ~ X1 * X2, data = df)
# wilcox.test()
```

