---
title: "Re-analyzing mobile phone usage at meal time (Moser et al., 2016)"
output: html_notebook
---
```{r setup}
if (!require("pacman")) install.packages("pacman", repos='https://stat.ethz.ch/CRAN/'); library(pacman)
p_load(ggplot2, 
       DT, 
       plyr,
       dplyr, 
       tidyr,
       assertthat,
       car,
       GetoptLong,
       tibble,
       cowplot,
       readr,
       broom,
       GGally)

# string interpolation
qq.options("code.pattern" = "#\\{CODE\\}") 

# plot theme
myTheme <- theme(panel.background = element_blank(), panel.grid.major = element_line(color="lightgrey", size = 0.2))

# decimal output
options(digits=2)

# datatable
options(DT.options = list(pageLength = 10))
options(DT.autoHideNavigation = TRUE)

# We ensure that the following packages are installed, but left unloaded to 
# prevent conflicts.
# (We can still use the functions inside, e.g., MASS::stepAIC() )

p_load(MASS); p_unload(MASS)     # for stepwise regression 
                
```

Moser et al. (2016) surveyed 1163 participants on their attitude of mobile phone usage at meal time. `meal_data` contains  the data they collected (which was published [online](http://labinthewild.org/data/index.php)). In the second part of their analysis, the explore what factors influence attitude about mobile phone behavior. In particular, they explore how the answers to the following questions:

(a) Age - `age`
(b) Do you own a mobile phone? (1 = yes, 2 = no)  — `question_11`
(c) What year did you get your first mobile phone? (number of years)  — `question_12`
(d) On a normal day, how often do you use your mobile phone?  — `question_16`
(e) How often do you use your mobile phone during meals with other people like friends and family?  — `question_14`
(f) How many individuals were included in the mealtime drawing? — Not included in the published dataset
(g) Child presence (0 = no, 1 = yes) — `child_included`
(h) Number of individuals that the respondent lives with - `housemates`
(i) Gender (0 = male, 1 = female, 2 = others) - `gender`


might influence the answers for the following questions:

* How appropriate would it be for an adult at this meal to take out their mobile phone and send or read a text message? — `question_6`
* How appropriate would it be for a child at this meal to take out their mobile phone and send or read a text message? — `question_7`
* How appropriate would it be for an adult at this meal to answer a mobile phone call? — `question_8`
* How appropriate would it be for an adult at this meal to take out their mobile phone and go online or use a social media site? — `question_9`
* How appropriate would it be for a child at this meal to take out their mobile phone and go online or use a social media site? — ` question_10`


**Note:** The dataset doesn't contain the responses to (f)

```{r}
# NOTE: the following data was preprocessed from the original Excel file by the script
#       ../data/TechAtMeals_CHI2016/preprocess.R
meal_data <- read_csv("../data/TechAtMeals_CHI2016/meal_data.csv" ) # actual data
meal_code <- read_csv("../data/TechAtMeals_CHI2016/meal_code.csv")  # dictionary of column names
```

## Automatic model comparison

We first try running automatic step AIC (which require removing rows with empty responses)
```{r}
lm_all <- lm(
  question_6 ~     # Response: Adult sending/reading texts
  age +            # (a) Age
  question_11 +    # (b) Do you own a mobile phone? (1 = yes, 2 = no)
  question_12 +    # (c) What year did you get your first mobile phone? (number of years)
  question_16 +    # (d) On a normal day, how often do you use your mobile phone?
  question_14 +    # (e) How often do you use your mobile phone during meals with other people like friends and family?
                   # (f) Missing: How many individuals were included in the mealtime drawing?
  child_included + # (g) Child presence (0 = no, 1 = yes)
  housemates +     # (h) Number of individuals that the respondent lives with
  gender,          # (i) Gender (0 = male, 1 = female, 2 = others)
  data = na.omit(meal_data)    # NEW: omit rows with empty response (this is different from the paper)
  )


step_backward_result <- 
  MASS::stepAIC(lm_all, 
    direction = "backward", 
    trace = TRUE           # this shows the fit trace so we can track AIC (optional)
    )

step_backward_result$anova
anova_result <- anova(step_backward_result)
print(anova_result)
```

The resulting model is quite different from the one in the paper. Looking at the `Residuals Df` in ANOVA table, we see `r anova_result$Df[4]`, which is substantially lower than the number of rows in the data (`r nrow(meal_data)` rows). Therefore, we have to check AIC manually.

## Manual model comparison

```{r}
lm_all <- lm(
  question_6 ~     # Response: Adult sending/reading texts
  age +            # (a) Age
  question_11 +    # (b) Do you own a mobile phone? (1 = yes, 2 = no)
  question_12 +    # (c) What year did you get your first mobile phone? (number of years)
  question_16 +    # (d) On a normal day, how often do you use your mobile phone?
  question_14 +    # (e) How often do you use your mobile phone during meals with other people like friends and family?
                   # (f) Missing: How many individuals were included in the mealtime drawing?
  child_included + # (g) Child presence (0 = no, 1 = yes)
  housemates +     # (h) Number of individuals that the respondent lives with
  gender,          # (i) Gender (0 = male, 1 = female, 2 = others)
  data = meal_data
  )

extractAIC(update(lm_all, . ~ .  - age))
extractAIC(update(lm_all, . ~ .  - question_11 ))
extractAIC(update(lm_all, . ~ .  - question_12 ))
extractAIC(update(lm_all, . ~ .  - question_16 ))
extractAIC(update(lm_all, . ~ .  - question_14 ))
extractAIC(update(lm_all, . ~ .  - child_included ))
extractAIC(update(lm_all, . ~ .  - hoursemates ))
extractAIC(update(lm_all, . ~ .  - gender ))  # lowest AIC: 1220


m2 <- update(lm_all, . ~ .  - gender )
extractAIC(update(m2, . ~ .  - age ))
extractAIC(update(m2, . ~ .  - question_11 ))
extractAIC(update(m2, . ~ .  - question_12 ))
extractAIC(update(m2, . ~ .  - question_16 ))
extractAIC(update(m2, . ~ .  - question_14 ))
extractAIC(update(m2, . ~ .  - child_included ))
extractAIC(update(m2, . ~ .  - hoursemates )) # lowest AIC:  1220

m3 <- update(m2, . ~ .  - housemates )
extractAIC(update(m3, . ~ .  - age ))
extractAIC(update(m3, . ~ .  - question_11 ))
extractAIC(update(m3, . ~ .  - question_12 ))
extractAIC(update(m3, . ~ .  - question_16 )) # lowest AIC: 1223 (which increased from the m3)
extractAIC(update(m3, . ~ .  - question_14 ))
extractAIC(update(m3, . ~ .  - child_included))

# therefore, we stop and choose m3
summary(m3)
```

Our results from model comparison still retain `question_11` and `question_12` as well as other coefficients.

## Final moodel from Moser et al. (2016)

```{r}
m_paper <- update(lm_all, . ~ . - question_11 - question_12 - housemates - gender, data = meal_data)
summary(m_paper)
```

Here, we have a reasonable residual Df of (`r summary(m_paper)$fstatistic[2]`), which is much closer to the number of rows in the data (`r nrow(meal_data)` rows). We also reproduced \\(F\\) and \\(R^2\\) from this model. There are possible explanation why our automatic AIC doesn't arrive at this model:

* Moser et al. (2016) removed `question_11` and `question_12` because their the t-test of these coefficients are not statistically significant.
* The missing (f) changes how the AIC develop in the model comparison sequence.

## Your turn
Try comparing models  with the second response variable: How appropriate would it be for a child at this meal to take out their mobile phone and send or read a text message? (`question_7`). Note that this variable is appropriate for those who indicated that they have children (`child_included` == 1).
```{r}
# TODO: your code here
```


# References

* Carol Moser, Sarita Y. Schoenebeck, and Katharina Reinecke. 2016. Technology at the Table: Attitudes about Mobile Phone Use at Mealtimes. In Proceedings of the 2016 CHI Conference on Human Factors in Computing Systems (CHI '16). ACM, New York, NY, USA, 1881-1892. DOI: https://doi.org/10.1145/2858036.2858357