---
title: "Repeated measures"
output: 
  html_notebook: 
    code_folding: hide
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)


####
# BEGIN: Chat's R header (revision: 10.4.17)
####

# The "pacman" package allows p_load() and p_load_gh() which
# automatically install missing packages and load them
if (!require("pacman")) install.packages("pacman", repos='https://stat.ethz.ch/CRAN/'); library(pacman)

p_load(
   car,        # grab bag of useful functions for NHST
   multcomp,   # for glht (planned comparison, post-hoc test)
   tidyverse,  # collection of the tidyverse packages (this automatically load the following):
   #dplyr,     #   - for data wrangling
   #tibble,    #   - a stricter alternative to data.frame
   #readr,     #   - a stricter alternative to read.csv
   #ggplot2,   #   - for plotting
               # other packages in tidyverse that are non-core
   stringr,    #   - for string functions
   tidyr,      #   - for data tidying
   forcats,    #   - utility functions for working with factor levels
               # extensions of tidyverse
   broom,      #   - for cleaing output from models, e.g., lm()
   cowplot,    #   - adds plot_grid() to put multiple ggplot()'s togeter
   GGally,     #   - adds ggpairs() which is a smarter scatterplot matrix
               # data structure
   GetoptLong, #   - string interpolation. See qq() explanation below
   lubridate,  #   - utility for parsing and performing arithematic on dates 
               # 
               # visualization & interactivity
   Hmisc,      #   - for plotting mean and CI in ggplot
   rafalib,    #   - for imagemat function (visualize contrast codings)
   DT,         #   - for showing data table with navigation/search controls
               # testing:
   assertthat  #   - unit-testing functions
)

p_load_gh(
   "eclarke/ggbeeswarm" # beeswarm plot extension for ggplot2
)

# GetoptLong config (string interpolation)
qq.options("code.pattern" = "#\\{CODE\\}") 

# ggplot2 config (plot theme)
myTheme <- theme(
   panel.background = element_blank(), 
   panel.grid.major = element_line(color="lightgrey", size = 0.2)
)

# DT config
options(DT.options = list(pageLength = 10))
options(DT.autoHideNavigation = TRUE)

# Optional: Decimal output readability
# options(digits=2)
##   NOTE: This option might cause output to be printed with rounding. (default value = 7)

####
# END: Chat's R header
####

p_load(lme4, lmerTest)  # for multilevel model functions
```

# Data and exploratory plot
The following dataset is simulated to be similar to the comparison of tangible knobs and touch input (Voelker et al., 2015). In this dataset, we simplify the study with the following setup:

* 2 Devices (Touch and Knob) x
* 3 Vision conditions (EyesFree, Peripheral, and EyesOn)
* Within-subjects design
* The dependent variable is the completion time in seconds.
```{r}
data1 <- tibble(
    Participant = rep(paste0("S", 1:8), 6), 
    Device = factor(
      c(rep("Touch",24), 
        rep("Knob",24)), levels = c("Touch", "Knob")), 
    Vision = factor(rep(
      c(rep("EyesOn",8),
        rep("Peripheral",8),
        rep("EyesFree",8)), 2), levels = c("EyesFree", "Peripheral", "EyesOn")),
    Time = c(1.2,1.4,1.8,2.0,1.1,1.5,1.5,1.7,
2.1,2.5,2.2,2.2,2.9,2.3,2.3,2.6,
3.5,3.4,3.3,3.2,2.9,2.8,3.8,3.4,
2.4,1.8,2.5,2.1,2.2,1.9,1.7,2.3,
2.8,3.1,3.2,4.0,2.9,3.6,3.2,3.7,
4.5,4.8,4.7,4.1,4.1,4.2,4.6,4.9))
data1
```

```{r}
# ensure that independent variables are factors (essential for glht())
data1$Device <- as.factor(data1$Device)
data1$Vision <- as.factor(data1$Vision)
data1$Participant <- as.factor(data1$Participant)
```

Plotting beeswarm of the data points and the mean time:

```{r}
pd <- position_dodge(0.3) # ensure no overlaps

plot_grid({ 
  data1 %>% 
    ggplot(aes(x = Device, y = Time, color = Vision)) +
    geom_beeswarm() + 
    expand_limits(y = 0) +
    theme(legend.position = "none")
  },{
  data1 %>% 
    ggplot(aes(x = Device, y = Time, color = Vision, group = Vision)) +
    stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3, position = pd) + 
    stat_summary(fun.y = mean, geom = "line", position = pd) + 
    stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0, position = pd) + 
    ylab("Mean(Time) and 95% CI") +
    expand_limits(y = 0)
  },
  nrow = 1)
```

# Model fitting
```{r}
# lmer() is from package lme4
m1 <- lmer(Time ~ (1|Participant) + Vision * Device, data = data1)
summary(m1)
anova(m1)
```

**Interpretation:** Both independent variables have statistically significant effect on the `Time`. The interaction effect is also statistically significant. We can follow up on this with either planned contrasts or pairwise comparison.


# References
* Simon Voelker, Kjell Ivar Øvergård, Chat Wacharamanotham, and Jan Borchers. 2015. Knobology Revisited: A Comparison of User Performance between Tangible and Virtual Rotary Knobs. In Proceedings of the 2015 International Conference on Interactive Tabletops & Surfaces (ITS '15). ACM, New York, NY, USA, 35-38. DOI: https://doi.org/10.1145/2817721.2817725