---
title: "Inference on one sample"
output: html_notebook
---

This notebook allows you to try inferencing. Feel free to make a copy of this notebook to try with other datasets.

```{r, include=FALSE}
# load required packages
library(ggplot2)  # for plotting graphs
library(DT)  # for showing data table
library(Hmisc) # for calculating mean and CI
```

We will analyze a synthetic data set from a hypothetical typing study. This study records the typing speed in words per minute (WPM), the error rate in percent, and the task completion time (TCT) in minutes.

We first load the data file and show it on a table.

```{r}
data <- read.csv("../data/typing-OneSample.csv")
datatable(data)
```

We can will plot the typing speed.

```{r}
ggplot(data, aes(x = WPM)) +
  geom_histogram(binwidth=.5)
```

Now, we calculate the mean and confidence intervals of the typing speed

```{r}
myStat <- mean_cl_normal(data$WPM)
names(myStat) <- c("Mean", "CILower", "CIUpper")  # rename the column so that they are self-explanatory
myStat
```

These statistics can be plotted below, together with a benchmark at 40 WPM.

```{r, fig.width = 8, fig.height = 1}
benchmark <- 40
ggplot(myStat ) +
  geom_point(aes(x = Mean), y = 0 ) + 
  geom_segment(aes(x = CILower, xend = CIUpper), y = 0, yend = 0 ) +
  geom_segment(aes(x = benchmark, xend = benchmark), y = -1, yend = 1 , colour = "lightgreen") +
  coord_fixed(ylim = c(-1, 1), ratio = 0.5)
```

Now we do a t-test against the benchmark.

```{r}
t.test(data$WPM, mu = benchmark)
```

Now, try plot and do t-test the error rate (benchmark 0) and the task completion time (benchmark 75) in the area below. (Press Ctrl + Enter or Cmd + Enter to run the code)

```{r}

```